---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cuda.ml

<!-- badges: start -->
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/cuda.ml)](https://cran.r-project.org/package=cuda.ml)
<a href="https://www.r-pkg.org/pkg/cuda.ml"><img src="https://cranlogs.r-pkg.org/badges/cuda.ml?color=brightgreen" style=""></a>
<!-- badges: end -->

The goal of {cuda.ml} is to provide a simple and intuitive R interface for
[RAPIDS cuML](https://github.com/rapidsai/cuml).
RAPIDS cuML is a suite of GPU-accelerated machine learning libraries powered by
[CUDA](https://en.wikipedia.org/wiki/CUDA).
{cuda.ml} is under active development, and currently implements R interfaces for
the algorithms listed below
(which is a subset of
[algorithms supported by RAPIDS cuML](https://github.com/rapidsai/cuml#supported-algorithms)).

### Supported Algorithms
| Category | Algorithm | Notes |
| --- | --- | --- |
| **Clustering** |  Density-Based Spatial Clustering of Applications with Noise (DBSCAN) | Only single-GPU implementation is supported at the moment |
|  | K-Means | Only single-GPU implementation is supported at the moment |
|  | Single-Linkage Agglomerative Clustering | |
| **Dimensionality Reduction** | Principal Components Analysis (PCA) | Only single-GPU implementation is supported at the moment |
| | Truncated Singular Value Decomposition (tSVD) | Only single-GPU implementation is supported at the moment |
| | Uniform Manifold Approximation and Projection (UMAP) | Only single-GPU implementation is supported at the moment |
| | t-Distributed Stochastic Neighbor Embedding (TSNE) | |
| **Linear Models for Regression or Classification** | Linear Regression (OLS) | |
| | Ridge, lasso, and elastic-net linear regression | |
| | Logistic and multinomial regression | |
| **Nonlinear Models for Regression or Classification** | Random Forest (RF) Classification | Training is single-GPU; inference uses nvForest. |
| | Random Forest (RF) Regression | Training is single-GPU; inference uses nvForest. |
| | nvForest inference for XGBoost, LightGBM, and Treelite models | CPU and GPU inference are supported. |
|  | K-Nearest Neighbors (KNN) Classification | Brute-force, IVFFlat, and IVFPQ indexes are supported. |
|  | K-Nearest Neighbors (KNN) Regression | Brute-force, IVFFlat, and IVFPQ indexes are supported. |
|  | Support Vector Machine Classifier (SVC) | |
|  | Epsilon-Support Vector Regression (SVR) | |

# Examples

## Using {cuda.ml} for supervised ML tasks through {parsnip}

{cuda.ml} provides {parsnip} bindings for supervised ML algorithms such as
`linear_reg`, `logistic_reg`, `multinom_reg`, `rand_forest`,
`nearest_neighbor`, `svm_rbf`, `svm_poly`, and `svm_linear`.

Regularized models follow tidymodels conventions for `penalty` and `mixture`.
When predictors need scaling, learn and apply it explicitly with a {recipes}
step such as `step_normalize(all_numeric_predictors())`.

The following example shows how {cuda.ml} can be used as a {parsnip} engine to
build a SVM classifier.

```{r svm-example, eval = FALSE}
library(dplyr, warn.conflicts = FALSE)
library(parsnip)
library(cuda.ml)
set.seed(11235)

train_inds <- iris %>%
  mutate(ind = row_number()) %>%
  group_by(Species) %>%
  slice_sample(prop = 0.7)

train_data <- iris[train_inds$ind, ]
test_data <- iris[-train_inds$ind, ]

model <- svm_rbf(mode = "classification", rbf_sigma = 10, cost = 50) %>%
  set_engine("cuda.ml") %>%
  fit(Species ~ ., data = train_data)

preds <- predict(model, test_data)

cat("Confusion matrix:\n\n")
preds %>%
  bind_cols(test_data %>% select(Species)) %>%
  yardstick::conf_mat(truth = Species, estimate = .pred_class)
```

## Using {cuda.ml} for unsupervised ML tasks

The following example shows how {cuda.ml} can be used for unsupervised ML tasks
such as k-means clustering.

```{r kmeans-example, eval = FALSE}
library(cuda.ml)

clustering <- cuda_ml_kmeans(
  iris[, which(names(iris) != "Species")],
  k = 3, max_iters = 100
)

# Expected outcome: there is strong correlation
# between cluster labels and `iris$Species`
print(clustering)

library(dplyr, warn.conflicts = FALSE)
tibble(cluster_id = clustering$labels, species = iris$Species) %>%
  group_by(cluster_id) %>% count(species)
```

## Using {cuda.ml} for visualizations

{cuda.ml} also features R interfaces for algorithms such as UMAP and t-SNE,
which are useful when one needs to visualize clusters of high-dimensional data
points by embedding them onto low-dimensional manifolds (i.e., 4 dimensions or
fewer).

For example, the code snippet below shows how `cuda_ml_umap()` can be used to
visualize the MNIST hand-written digits dataset, and also, the coloring based on
the true label of each sample demonstrates how well the UMAP algorithm
transforms different hand writings of the same digit into nearby points in a 2D
embedding:

```{r umap-example, eval = FALSE}
library(cuda.ml)
library(ggplot2)
library(magrittr)

# load mnist
source("data-raw/load-mnist.R")
str(mnist_images)
str(mnist_labels)


# flatten each image to a 1d array, combine into a matrix with 1 row per image
flatten <- function(img) {
  dim(img) <- NULL
  img
}

flattened_mnist_images <-
  mnist_images %>% asplit(3) %>% lapply(flatten) %>% do.call(rbind, .)

# embed
embedding <- cuda_ml_umap(
  flattened_mnist_images, n_components = 2, n_neighbors = 50,
  local_connectivity = 15, repulsion_strength = 10
)

str(embedding$transformed_data)

# visualize
embedding$transformed_data %>%
  as.data.frame() %>%
  dplyr::mutate(Label = factor(mnist_labels)) %>%
  ggplot(aes(x = V1, y = V2, color = Label)) +
  geom_point(alpha = .5, size = .5) +
  labs(title = "UMAP: Uniform Manifold Approximation and Projection",
       subtitle = "Two Dimensional Embedding of MNIST")

```

From this type of visualization, we can qualitatively understand the following
about the MNIST dataset:

- The dataset can be reasonably classified into some number of categories.
- The right number of categories may be any where between 9 and 11.
- While there are some categories that are clearly distinguishable from others,
  there are others that have less clear boundaries with their neighbors.
- A small fraction of data points did not fit particularly well into any of the
  categories.
- Most data points belonging to the same digit category are clustered together
  in the UMAP output


## Installation

Install the R package from CRAN, then prepare its native backend and runtime:

``` r
install.packages("cuda.ml")
cuda.ml::cuda_ml_install()
```

The CRAN package is a portable R installer and loader. It contains no compiled
code, so `install.packages("cuda.ml")` does not need a compiler, CUDA, RAPIDS,
Python, or conda. Loading it is silent and side-effect free:

``` r
library(cuda.ml)
info <- cuda_ml_backend_info()
stopifnot(
  identical(info$backend, "download"),
  info$backend_available
)
```

`library(cuda.ml)` does not inspect the GPU, create a cache, contact the network,
or load native code. `cuda_ml_backend_info()` reports the selected platform and
R-version backend, its exact library versions, and whether the managed cache is
complete. It does not report whether a GPU can execute a model.

### Runtime provisioning

`cuda_ml_install()` first downloads the small backend archive for the current R
minor version from the package's GitHub Releases. It verifies the archive and
native library against hashes shipped in the R package. It then downloads the
locked CUDA 13.2.2 and RAPIDS cuML and nvForest 26.06 wheels directly from their
upstream Python package hosts. The current runtime is about 1.6 GiB, so this step
can take several minutes. Treelite 4.7.0 is linked into the backend and is not a
runtime download.

`cuda_ml_install()` does not require a GPU or NVIDIA driver, and repeated calls
reuse the completed cache. It does not load the backend or initialize CUDA.
Model operations do not provision the runtime implicitly; when the cache is
absent, they report the installation command. After provisioning, GPU-backed
operations require only a supported NVIDIA GPU and driver. nvForest CPU
inference does not require a GPU or driver.

The default cache is `tools::R_user_dir("cuda.ml", "cache")`. Set
`CUDA_ML_CACHE_DIR` to use a different location:

``` r
Sys.setenv(CUDA_ML_CACHE_DIR = "/opt/cuda-ml-cache")
cuda.ml::cuda_ml_install()
```

For an internal or offline mirror, set `CUDA_ML_BACKEND_MIRROR` to an `https://`
or `file://` directory containing the exact locked backend archive. Hash
verification remains enabled.

### Supported systems

Prebuilt backends target Linux x86_64 with glibc 2.28 or newer rather than a
specific distribution. This includes current Ubuntu, Debian, RHEL-compatible,
and WSL2 Linux distributions that meet the glibc requirement.
Following the
[RAPIDS 26.06 platform requirements](https://docs.rapids.ai/platform-support/),
the binary requires an NVIDIA driver version 580 or newer and supports GPU
compute capabilities 7.5, 8.0, 8.6, 8.9, 9.0, 10.0, and 12.0. The compute
capability 12.0 PTX image also provides forward compatibility for newer GPUs
supported by CUDA, following
[CUDA's forward-compatibility model](https://docs.nvidia.com/cuda/archive/13.2.0/cuda-compiler-driver-nvcc/index.html).

Native Windows, macOS, Linux ARM64, musl-based Linux distributions, and glibc
versions older than 2.28 are not currently supported.

### Packaging and source

CRAN installation and checks are network-free. The package contacts the network
only when `cuda_ml_install()` is called explicitly. Native backend archives are
built in a pinned manylinux 2.28 container, audited for their glibc and
libstdc++ requirements, and hosted as GitHub Release assets. One archive is
published for each supported R minor version.

The native source remains in the source package for advanced local builds. Set
`CUDA_ML_BUILD_MODE=local`, supply CUDA Toolkit 13.2.2 through `CUDA_HOME`, and
supply a prefix through `CUML_PREFIX` containing cuML and nvForest 26.06,
Treelite 4.7.0 headers, and `lib/libtreelite_static.a`. Build Treelite as
position-independent code with its default libstdc++ ABI and with OpenMP
disabled. Local builds never download or provision Treelite. Set
`CUML_CUDA_ARCHITECTURES` explicitly to the CMake CUDA architectures to
compile, and set `CUDA_ML_CXX` to GNU C++ 14 or newer. The same compiler is used
for C++ sources and nvcc host compilation. Missing inputs and other library
versions fail at configuration time.

### Development version

A development installation uses the same downloaded backend pathway:

``` r
# install.packages("devtools")
devtools::install_github("mlverse/cuda.ml")
cuda.ml::cuda_ml_install()
```

## Appendix

<details>
<summary>Inspect MNIST images</summary>
```{r mnist, eval = FALSE}
plot_mnist(1:64)
```
</details>
