---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cuda.ml

<!-- badges: start -->
[![CRAN\_Status\_Badge](https://www.r-pkg.org/badges/version/cuda.ml)](https://cran.r-project.org/package=cuda.ml)
<a href="https://www.r-pkg.org/pkg/cuda.ml"><img src="https://cranlogs.r-pkg.org/badges/cuda.ml?color=brightgreen" style=""></a>
<!-- badges: end -->

The goal of {cuda.ml} is to provide a simple and intuitive R interface for
[RAPIDS cuML](https://github.com/rapidsai/cuml).
RAPIDS cuML is a suite of GPU-accelerated machine learning libraries powered by
[CUDA](https://en.wikipedia.org/wiki/CUDA).
{cuda.ml} is under active development, and currently implements R interfaces for
the algorithms listed below
(which is a subset of
[algorithms supported by RAPIDS cuML](https://github.com/rapidsai/cuml#supported-algorithms)).

### Supported Algorithms
| Category | Algorithm | Notes |
| --- | --- | --- |
| **Clustering** |  Density-Based Spatial Clustering of Applications with Noise (DBSCAN) | Only single-GPU implementation is supported at the moment |
|  | K-Means | Only single-GPU implementation is supported at the moment |
|  | Single-Linkage Agglomerative Clustering | |
| **Dimensionality Reduction** | Principal Components Analysis (PCA) | Only single-GPU implementation is supported at the moment |
| | Truncated Singular Value Decomposition (tSVD) | Only single-GPU implementation is supported at the moment |
| | Uniform Manifold Approximation and Projection (UMAP) | Only single-GPU implementation is supported at the moment |
| | Random Projection | |
| | t-Distributed Stochastic Neighbor Embedding (TSNE) | |
| **Linear Models for Regression or Classification** | Linear Regression (OLS) | |
| | Linear Regression with Lasso or Ridge Regularization | |
| | Logistic Regression | |
| **Nonlinear Models for Regression or Classification** | Random Forest (RF) Classification | Only single-GPU implementation is supported at the moment |
| | Random Forest (RF) Regression | Only single-GPU implementation is supported at the moment |
| | Inference for decision tree-based models in XGBoost or LightGBM formats using the CuML Forest Inference Library (FIL) | Requires linkage to the Treelite C library when {cuml} is installed. Treelite is used for model loading. |
|  | K-Nearest Neighbors (KNN) Classification | Uses [Faiss](https://github.com/facebookresearch/faiss) for Nearest Neighbors Query. |
|  | K-Nearest Neighbors (KNN) Regression | Uses [Faiss](https://github.com/facebookresearch/faiss) for Nearest Neighbors Query. |
|  | Support Vector Machine Classifier (SVC) | |
|  | Epsilon-Support Vector Regression (SVR) | |

# Examples

## Using {cuda.ml} for supervised ML tasks through {parsnip}

{cuda.ml} provides {parsnip} bindings for supervised ML algorithms such as
`rand_forest`, `nearest_neighbor`, `svm_rbf`, `svm_poly`, and `svm_linear`.

The following example shows how {cuda.ml} can be used as a {parsnip} engine to
build a SVM classifier.

```{r svm-example, eval = FALSE}
library(dplyr, warn.conflicts = FALSE)
library(parsnip)
library(cuda.ml)
set.seed(11235)

train_inds <- iris %>%
  mutate(ind = row_number()) %>%
  group_by(Species) %>%
  slice_sample(prop = 0.7)

train_data <- iris[train_inds$ind, ]
test_data <- iris[-train_inds$ind, ]

model <- svm_rbf(mode = "classification", rbf_sigma = 10, cost = 50) %>%
  set_engine("cuda.ml") %>%
  fit(Species ~ ., data = train_data)

preds <- predict(model, test_data)

cat("Confusion matrix:\n\n")
preds %>%
  bind_cols(test_data %>% select(Species)) %>%
  yardstick::conf_mat(truth = Species, estimate = .pred_class)
```

## Using {cuda.ml} for unsupervised ML tasks

The following example shows how {cuda.ml} can be used for unsupervised ML tasks
such as k-means clustering.

```{r kmeans-example, eval = FALSE}
library(cuda.ml)

clustering <- cuda_ml_kmeans(
  iris[, which(names(iris) != "Species")],
  k = 3, max_iters = 100
)

# Expected outcome: there is strong correlation
# between cluster labels and `iris$Species`
print(clustering)

library(dplyr, warn.conflicts = FALSE)
tibble(cluster_id = clustering$labels, species = iris$Species) %>%
  group_by(cluster_id) %>% count(species)
```

## Using {cuda.ml} for visualizations

{cuda.ml} also features R interfaces for algorithms such as UMAP and t-SNE,
which are useful when one needs to visualize clusters of high-dimensional data
points by embedding them onto low-dimensional manifolds (i.e., 4 dimensions or
fewer).

For example, the code snippet below shows how `cuda_ml_umap()` can be used to
visualize the MNIST hand-written digits dataset, and also, the coloring based on
the true label of each sample demonstrates how well the UMAP algorithm
transforms different hand writings of the same digit into nearby points in a 2D
embedding:

```{r umap-example, eval = FALSE}
library(cuda.ml)
library(ggplot2)
library(magrittr)

# load mnist
source("data-raw/load-mnist.R")
str(mnist_images)
str(mnist_labels)


# flatten each image to a 1d array, combine into a matrix with 1 row per image
flatten <- function(img) {
  dim(img) <- NULL
  img
}

flattened_mnist_images <-
  mnist_images %>% asplit(3) %>% lapply(flatten) %>% do.call(rbind, .)

# embed
embedding <- cuda_ml_umap(
  flattened_mnist_images, n_components = 2, n_neighbors = 50,
  local_connectivity = 15, repulsion_strength = 10
)

str(embedding$transformed_data)

# visualize
embedding$transformed_data %>%
  as.data.frame() %>%
  dplyr::mutate(Label = factor(mnist_labels)) %>%
  ggplot(aes(x = V1, y = V2, color = Label)) +
  geom_point(alpha = .5, size = .5) +
  labs(title = "UMAP: Uniform Manifold Approximation and Projection",
       subtitle = "Two Dimensional Embedding of MNIST")

```

From this type of visualization, we can qualitatively understand the following
about the MNIST dataset:

- The dataset can be reasonably classified into some number of categories.
- The right number of categories may be any where between 9 and 11.
- While there are some categories that are clearly distinguishable from others,
  there are others that have less clear boundaries with their neighbors.
- A small fraction of data points did not fit particularly well into any of the
  categories.
- Most data points belonging to the same digit category are clustered together
  in the UMAP output


## Installation

### R-universe binary

The R-universe binary is the supported no-compiler installation for its current
Linux target: Ubuntu 26.04 (Resolute) x86_64, including WSL2 running that
distribution. Use
[R-universe's Linux binary repository](https://docs.r-universe.dev/install/binaries.html)
rather than its source-package repository:

``` r
linux_binary_repo <- function(universe) {
  r_version <- paste(
    R.version$major,
    strsplit(R.version$minor, ".", fixed = TRUE)[[1L]][1L],
    sep = "."
  )
  sprintf(
    "https://%s.r-universe.dev/bin/linux/resolute-%s/%s/",
    universe,
    R.version$arch,
    r_version
  )
}

repos <- c(
  mlverse = linux_binary_repo("mlverse"),
  CRAN = linux_binary_repo("cran")
)
stopifnot(
  identical(unname(Sys.info()[["sysname"]]), "Linux"),
  identical(R.version$arch, "x86_64"),
  grepl(
    "/bin/linux/resolute-x86_64/[0-9]+[.][0-9]+/$",
    repos[["mlverse"]]
  )
)

install.packages(
  "cuda.ml",
  repos = repos
)
```

The binary repository path selects the prebuilt tarball. Stock Linux R does not
support `type = "binary"`, so leave `type` at its default. The `stopifnot()`
check prevents accidentally installing from the source endpoint or an
unsupported architecture.

The binary contains a precompiled {cuda.ml} backend, but not the CUDA and RAPIDS
runtime libraries. Loading the package is silent and side-effect free:

``` r
library(cuda.ml)
stopifnot(has_cuML())
```

`library(cuda.ml)` does not inspect the GPU, create a cache, contact the network,
or load the native backend. `has_cuML()` reports whether the installed package
contains a functional backend; it does not report whether a runtime or GPU is
ready. The assertion also catches an unavailable binary that fell back to the
CRAN-compatible source stub.

### Runtime provisioning

The first cuML operation downloads and verifies the exact CUDA 13.2 and RAPIDS
26.06 runtime required by the binary. The current lock downloads about 1.6 GiB,
so the first call can take several minutes. To prewarm a container or machine
instead, run:

``` r
cuda.ml::cuda_ml_install()
```

`cuda_ml_install()` does not require a GPU or NVIDIA driver, and repeated calls
reuse the completed cache. It does not load the backend or initialize CUDA.
After provisioning, the target machine only needs a supported NVIDIA GPU and
driver.

The default cache is `tools::R_user_dir("cuda.ml", "cache")`. Set
`CUDA_ML_CACHE_DIR` to use a different location:

``` r
Sys.setenv(CUDA_ML_CACHE_DIR = "/opt/cuda-ml-cache")
cuda.ml::cuda_ml_install()
```

### Supported systems

R-universe currently publishes the managed binary for Ubuntu 26.04 (Resolute)
x86_64. This includes WSL2 when its Linux distribution is Ubuntu 26.04.
Following the
[RAPIDS 26.06 platform requirements](https://docs.rapids.ai/platform-support/),
the binary requires an NVIDIA driver version 580 or newer and supports GPU
compute capabilities 7.5, 8.0, 8.6, 8.9, 9.0, 10.0, and 12.0. The compute
capability 12.0 PTX image also provides forward compatibility for newer GPUs
supported by CUDA, following
[CUDA's forward-compatibility model](https://docs.nvidia.com/cuda/archive/13.2.0/cuda-compiler-driver-nvcc/index.html).

Other Linux distributions, native Windows, macOS, and Linux ARM64 are not yet
supported by the managed binary.

### CRAN and source builds

CRAN builds are network-free and install an explicit stub backend.
`has_cuML()` returns `FALSE` for that build, and `cuda_ml_install()` directs
users to the R-universe binary. Install from R-universe when a functional
binary without local compilation is required.

Advanced local source builds remain available. Supply an explicit CUDA 13.2
toolkit through `CUDA_HOME` and a RAPIDS 26.06 prefix through `CUML_PREFIX`.
Other CUDA or RAPIDS versions do not match the managed runtime and are rejected.

### Development version

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mlverse/cuda.ml")
```

To speed up recompilation times during development, set this env var:

```bash
echo "export CUML4R_ENABLE_CCACHE=1" >> ~/.bashrc
. ~/.bashrc
```


## Appendix

<details>
<summary>Inspect MNIST images</summary>
```{r mnist, eval = FALSE}
plot_mnist(1:64)
```
</details>
