---
title: "Install and manage cuda.ml"
description: >
  Install a prebuilt or source-built backend, inspect and audit its cache, and
  prepare GPU or CPU-only deployments.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Install and manage cuda.ml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The cuda.ml R package contains no compiled code. Installing or loading the R
package does not download a native backend, inspect a GPU, create a cache, or
initialize CUDA. Provision the backend explicitly for each environment before
running native operations.

The examples in this vignette are not evaluated when it is built, so building
the package does not require a network connection, a native backend, or a GPU.

## Choose a backend

cuda.ml provides two prebuilt deployment paths:

| Need | Install | Contents and requirements |
|---|---|---|
| Train cuML models or run GPU inference | `cuda_ml_install()` | Complete managed CUDA, RAPIDS cuML, and nvForest runtime; GPU operations require a supported NVIDIA GPU and driver. |
| Run nvForest inference only on CPU | `cuda_ml_install(device = "cpu")` | Separate CPU backend with no cuML or CUDA runtime libraries; no NVIDIA GPU or driver is required. |

The complete backend is currently about 1.6 GiB. The CPU-only nvForest backend
is roughly 1 MiB to download and 3 MiB when installed. It can load, restore,
and run nvForest models, including random forests trained on a GPU with
`cuda_ml_rand_forest()`, but it cannot train cuML models or run GPU inference.

An installation of the complete backend can also execute nvForest models on
CPU. The smaller backend is intended for deployments that do not otherwise
need the managed CUDA and RAPIDS runtime. See
[nvForest inference and deployment](nvforest.html) for training, import,
persistence, and deployment workflows.

## Install the default prebuilt backend

Install the R package, then provision the complete backend:

```{r install-gpu}
install.packages("cuda.ml")

library(cuda.ml)
cuda_ml_install()
```

`cuda_ml_install()` downloads the prebuilt backend selected for the current R
minor version and its exact locked runtime libraries. The installer verifies
the downloaded files and stores the completed installation in the cuda.ml user
cache. It does not require a GPU or NVIDIA driver, load the backend, or
initialize CUDA.

Calls with the same inputs reuse the completed cache and are a no-op. Model
loading and prediction never install a missing backend implicitly; their error
messages identify the required installation call.

For a first model workflow after installation, continue with [Getting started
with cuda.ml](cuda-ml.html).

## Install CPU-only nvForest inference

On a host that will only run nvForest models on CPU, install the separate
backend:

```{r install-cpu}
library(cuda.ml)
cuda_ml_install(device = "cpu")
```

CPU-only installation is prebuilt only. `source = TRUE`, `dependencies`, and
`architectures` do not apply to this path. Installation remains explicit: a
CPU prediction does not download this backend when it is absent.

## Supported systems

The native backends require Linux x86_64 with glibc 2.28 or newer. This covers
current Ubuntu, Debian, RHEL-compatible, and WSL2 Linux distributions that meet
the glibc requirement. Native Windows, macOS, Linux ARM64, musl-based Linux,
and systems with an older glibc are not supported.

Installing the complete backend does not require a GPU. GPU-backed operations
require NVIDIA driver 580 or newer and a supported GPU. The prebuilt backend
contains real targets for compute capabilities 7.5, 8.0, 8.6, 8.9, 9.0, 10.0,
and 12.0, plus the compute capability 12.0 PTX image for CUDA forward
compatibility. CPU-only nvForest inference has the same operating-system
requirement but requires neither an NVIDIA GPU nor an NVIDIA driver.

## Place or mirror the cache

By default, cuda.ml uses:

```{r default-cache}
tools::R_user_dir("cuda.ml", "cache")
```

Set `CUDA_ML_CACHE_DIR` before installation to put managed files elsewhere:

```{r custom-cache}
Sys.setenv(CUDA_ML_CACHE_DIR = "/opt/cuda-ml-cache")
cuda_ml_install()
```

Set `CUDA_ML_BACKEND_MIRROR` to an `https://` or `file://` directory that
contains the exact locked backend archive:

```{r backend-mirror}
Sys.setenv(
  CUDA_ML_BACKEND_MIRROR = "file:///srv/cuda-ml-backends"
)
cuda_ml_install()
```

The installer appends the locked archive filename to that directory and keeps
hash verification enabled. Other URL schemes are rejected.

For the complete backend, this setting mirrors only the cuda.ml backend
archive. Managed CUDA and RAPIDS runtime artifacts still use the locked
upstream URLs recorded in the package manifest.

## Inspect the selected backend

`cuda_ml_backend_info()` performs read-only cache and inventory checks:

```{r backend-info}
info <- cuda_ml_backend_info()

info[c(
  "backend",
  "build_mode",
  "runtime_installed",
  "runtime_path",
  "backend_loaded"
)]

info[c(
  "nvforest_cpu_runtime_installed",
  "nvforest_cpu_runtime_path",
  "nvforest_cpu_backend_loaded"
)]
```

It reports the selected backend, locked library versions, supported
architectures, cache status, and the separate CPU-only nvForest status. It does
not create or change the cache, use the network, inspect an NVIDIA GPU or
driver, or load native code. Consequently, it reports whether the exact cache
is complete, not whether a GPU can execute a model.

## Audit or remove cached backends

Ordinary cache reuse checks markers, inventories, sizes, and links. Run an
explicit audit when you need to recompute the recorded hashes and validate
native registration:

```{r audit}
cuda_ml_runtime_audit()
cuda_ml_runtime_audit(device = "cpu")
```

For the complete downloaded backend, the audit also validates the managed
runtime dependency closure. The default audits the selected complete backend;
if a source build is selected, it audits that source-built backend.

To remove downloaded, source-built, and CPU-only backend cache generations,
including the selected-backend record, start a fresh R session and run:

```{r clean-cache}
library(cuda.ml)
cuda_ml_cache_clean()
```

Restart R first if either native backend has been loaded. After cleaning,
provision the backend needed by the environment again.

## Build the complete backend from source

Source installation builds the complete GPU-capable backend. It does not build
the separate CPU-only backend.

### Managed dependencies

The default source path downloads no prebuilt cuda.ml backend. It downloads and
verifies the locked CUDA 13.2.2 and RAPIDS 26.06 development artifacts, CMake,
and Ninja, builds Treelite 4.7.0 statically, and caches the toolchain:

```{r source-managed}
cuda_ml_install(source = TRUE)
```

The managed build requires Linux x86_64 with glibc 2.28 or newer and GNU C++ 14
or newer. It does not require Python, Conda, Docker, a system CUDA Toolkit, a
system RAPIDS installation, or a GPU. If `CUDA_ML_CXX` is unset, the installer
prefers `g++-14` and then `g++` on `PATH`.

By default, the build uses `nvidia-smi` to detect distinct CUDA-visible compute
capabilities and compiles their real targets. It honors
`CUDA_VISIBLE_DEVICES`. If detection is unavailable, it uses cuda.ml's portable
architecture list. Choose the target policy explicitly when needed:

```{r source-architectures}
# Require successful detection from a CUDA-visible GPU.
cuda_ml_install(source = TRUE, architectures = "native")

# Build the package's relocatable architecture set.
cuda_ml_install(source = TRUE, architectures = "portable")

# Build explicit CMake CUDA targets.
cuda_ml_install(
  source = TRUE,
  architectures = "86-real;89-real"
)
```

A native-target build usually takes less time and space, but the resulting
backend supports only the selected architectures. Repeated calls with the same
build inputs reuse the cached toolchain and backend. Restart R before changing
the selected prebuilt or source-built backend, or source-build inputs, after a
native backend has been loaded in that process.

### Host dependencies

The host path makes no downloads and requires every build input explicitly:

```{r source-host}
Sys.setenv(
  CUDA_HOME = "/usr/local/cuda-13.2",
  CUML_PREFIX = "/opt/rapids-26.06",
  CUML_CUDA_ARCHITECTURES = "86-real",
  CUDA_ML_CXX = "/usr/bin/g++-14"
)

cuda_ml_install(source = TRUE, dependencies = "host")
```

`CUDA_HOME` must contain the locked CUDA Toolkit 13.2.2. `CUML_PREFIX` must
contain cuML and nvForest 26.06, Treelite 4.7.0 headers, and
`lib/libtreelite_static.a`. GNU C++ 14 or newer is required, and CMake 3.21.1
or newer must be on `PATH`. The CUDA, RAPIDS, and Treelite prefixes must remain
in place because the compiled backend links to their libraries.

## Troubleshooting

| Symptom | Action |
|---|---|
| A native model operation says the managed runtime is not installed. | Run `cuda_ml_install()` once in that environment. |
| CPU inference says the CPU-only nvForest backend is not installed. | Run `cuda_ml_install(device = "cpu")`. |
| An operation says cuda.ml requires Linux x86_64 with glibc 2.28 or newer. | Move the native operation to a supported system; the CPU-only backend has the same platform requirement. |
| An audit reports failed content validation. | Start a fresh R session, run `cuda_ml_cache_clean()`, then reinstall the complete or CPU-only backend. |
| Cleaning the cache or changing the selected backend asks for a restart. | Restart R so no backend is loaded, then repeat the operation. |
| A managed source build cannot find a suitable compiler. | Install GNU C++ 14 or newer and, if needed, set `CUDA_ML_CXX` to its executable. |
| `architectures = "native"` cannot detect a GPU. | Make a GPU and `nvidia-smi` visible, use `architectures = "portable"`, or supply explicit targets such as `"86-real"`. |
| A host source build reports missing inputs. | Set `CUDA_HOME`, `CUML_PREFIX`, `CUML_CUDA_ARCHITECTURES`, and `CUDA_ML_CXX` to installations with the locked versions. |
