---
title: "Save and restore models"
description: >
  Choose a persistence format and move fitted cuda.ml models between R
  sessions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Save and restore models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") ||
    identical(Sys.getenv("CUDA_ML_GPU_VIGNETTES"), "true")
)
```

Some fitted cuda.ml objects contain native pointers that are meaningful only in
the R process that created them. Use an explicit cuda.ml model state, a
`bundle` object, or an nvForest checkpoint pair instead of relying on
`saveRDS()` to capture a live fitted object.

## Choose a persistence format

| Need | Save | Restore | Result |
|:---|:---|:---|:---|
| One R-native artifact | `cuda_ml_serialize()` | `cuda_ml_unserialize()` | A cuda.ml model state |
| Integration with the `bundle` package | `bundle::bundle()` and `saveRDS()` | `readRDS()` and `bundle::unbundle()` | A bundle containing the same cuda.ml state |
| A Treelite checkpoint usable outside cuda.ml | `cuda_ml_nvforest_export()` | `cuda_ml_nvforest_import()` | A checkpoint plus cuda.ml metadata |

The first two choices support cuda.ml models that implement explicit model
state. The checkpoint choice is only for random forests and other
nvForest-backed models. It is described in more detail in the
[nvForest inference guide](nvforest.html).

## Save directly to a file

The simplest file workflow passes a path directly. cuda.ml writes a
gzip-compressed model state rather than a live native pointer.

```{r save-file}
library(cuda.ml)

cuda_ml_install()

model <- cuda_ml_linear_reg(
  mpg ~ .,
  data = mtcars,
  penalty = 0.01,
  mixture = 0
)

state_path <- tempfile(fileext = ".cuda-ml-state")
cuda_ml_serialize(model, state_path)
```

In a new R process, prepare the required backend and restore the model. cuda.ml
validates the state and backend before loading it.

```{r restore-file}
library(cuda.ml)

cuda_ml_install()

model <- cuda_ml_unserialize(state_path)

predict(model, mtcars[1:5, names(mtcars) != "mpg"])
```

## Keep the state as raw bytes

With its default `connection = NULL`, `cuda_ml_serialize()` returns the
uncompressed state as a raw vector. This is useful for object stores and other
systems that accept bytes directly.

```{r raw-state}
state <- cuda_ml_serialize(model)
str(state)

model <- cuda_ml_unserialize(state)
```

The `blob` package can wrap this raw vector as one database BLOB value.

## Use a bundle

The `bundle` package wraps the same explicit cuda.ml state and records how to
restore it. This is useful in workflows that already use `bundle`.

```{r bundle}
library(bundle)

bundle_path <- tempfile(fileext = ".bundle.rds")
bundled_model <- bundle(model)
saveRDS(bundled_model, bundle_path)

bundled_model <- readRDS(bundle_path)
model <- unbundle(bundled_model)
```

For an nvForest-backed model, `device` chooses where the bundle will restore.
Use this when training a random forest on a GPU and deploying it on a CPU-only
host.

```{r bundle-nvforest}
forest <- cuda_ml_rand_forest(
  Species ~ .,
  data = iris,
  trees = 100,
  seed = 1
)

cpu_bundle <- bundle(forest, device = "cpu")
forest_bundle_path <- tempfile(fileext = ".bundle.rds")
saveRDS(cpu_bundle, forest_bundle_path)
```

The target host must prepare a backend that supports CPU inference before
calling `unbundle()`. For a smaller CPU-only deployment:

```{r unbundle-nvforest}
library(cuda.ml)
library(bundle)

cuda_ml_install(device = "cpu")
forest <- unbundle(readRDS(forest_bundle_path))
```

The `device` argument is supported only for nvForest-backed models.

## Export an nvForest checkpoint pair

`cuda_ml_nvforest_export()` writes two files:

- `<prefix>.treelite.checkpoint`, containing the device-neutral trees;
- `<prefix>.cuda-ml.json`, containing the metadata needed for an exact cuda.ml
  round-trip.

```{r export-nvforest}
forest_directory <- tempfile("forest-artifact-")
dir.create(forest_directory)
cuda_ml_nvforest_export(
  forest,
  directory = forest_directory,
  prefix = "model"
)
```

Copy both files when another cuda.ml process will restore the model. Select the
deployment device during import:

```{r import-nvforest}
cuda_ml_install(device = "cpu")

forest <- cuda_ml_nvforest_import(
  directory = forest_directory,
  prefix = "model",
  device = "cpu"
)
```

Other Treelite consumers can read the checkpoint alone, but they must supply
predictors in the recorded order and reproduce any class-label and
postprocessing semantics in the JSON sidecar. Loading the bare checkpoint back
into cuda.ml does not recover those semantics. Use the pair for an exact
round-trip.

## Select the nvForest restore device

Current random-forest and nvForest states contain device-neutral Treelite model
bytes. Select CPU or GPU inference while restoring:

```{r nvforest-devices}
forest_state <- cuda_ml_serialize(forest)
cpu_forest <- cuda_ml_unserialize(forest_state, device = "cpu")
gpu_forest <- cuda_ml_unserialize(
  forest_state,
  device = "gpu",
  device_id = 0
)
```

If `device` is omitted, these states restore for GPU inference. The state keeps
prediction precision, class labels, preprocessing, and model semantics. It does
not keep the deployment device, device identifier, tree layout, chunk size, or
memory alignment. Restore-time inference options are supported only for current
nvForest and random-forest states.

Install the complete backend for GPU operation or the smaller CPU backend for
CPU-only nvForest inference. See the
[installation and runtime guide](install-manage.html) for those workflows.

## Treat model artifacts as trusted input

Load cuda.ml states, bundles, and checkpoint sidecars only from trusted
sources. An nvForest JSON sidecar embeds an R-serialized preprocessing
blueprint. Its SHA-256 digest checks checkpoint integrity, not the identity of
the artifact's producer.

```{r cleanup, include = FALSE}
unlink(
  c(state_path, bundle_path, forest_bundle_path, forest_directory),
  recursive = TRUE
)
```
