---
title: "Save and restore models"
description: >
  Choose a persistence format, move fitted cuda.ml models between R sessions,
  and understand backend compatibility requirements.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Save and restore models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Some fitted cuda.ml objects contain native pointers that are meaningful only in
the R process that created them. Use an explicit cuda.ml model state, a
`bundle` object, or an nvForest checkpoint pair instead of relying on
`saveRDS()` to capture a live fitted object.

The examples are not evaluated when this vignette is built. Building the
vignette therefore requires no GPU, native backend, runtime download, or
network access.

## Choose a persistence format

| Need | Save | Restore | Result |
|:---|:---|:---|:---|
| One R-native artifact | `cuda_ml_serialize()` | `cuda_ml_unserialize()` | A versioned cuda.ml model state |
| Integration with the `bundle` package | `bundle::bundle()` and `saveRDS()` | `readRDS()` and `bundle::unbundle()` | A bundle containing the same cuda.ml state |
| A Treelite checkpoint usable outside cuda.ml | `cuda_ml_nvforest_export()` | `cuda_ml_nvforest_import()` | A checkpoint plus cuda.ml metadata |

The first two choices support cuda.ml models that implement explicit model
state. The checkpoint choice is only for random forests and other
nvForest-backed models. It is described in more detail in the
[nvForest inference guide](nvforest.html).

## Save one R-native state

`cuda_ml_serialize()` returns a raw vector when `connection = NULL`, its
default. The vector contains a versioned model state rather than a live native
pointer.

```{r save-state}
library(cuda.ml)

cuda_ml_install()

model <- cuda_ml_linear_reg(
  mpg ~ .,
  data = mtcars,
  penalty = 0.01,
  mixture = 0
)

state <- cuda_ml_serialize(model)
saveRDS(state, "mtcars-ridge.cuda-ml-state.rds")
```

In a new R process, prepare the required backend before using the restored
model, then pass the saved raw vector to `cuda_ml_unserialize()`.

```{r restore-state}
library(cuda.ml)

cuda_ml_install()

state <- readRDS("mtcars-ridge.cuda-ml-state.rds")
model <- cuda_ml_unserialize(state)
predict(model, mtcars[1:5, names(mtcars) != "mpg"])
```

You can write the state directly to a binary connection instead:

```{r connection}
connection <- file("mtcars-ridge.cuda-ml-state", open = "wb")
cuda_ml_serialize(model, connection)
close(connection)

connection <- file("mtcars-ridge.cuda-ml-state", open = "rb")
model <- cuda_ml_unserialize(connection)
close(connection)
```

## Use a bundle

The `bundle` package wraps the same explicit cuda.ml state and records how to
restore it. This is useful in workflows that already use `bundle`; it does not
change cuda.ml's compatibility requirements.

```{r bundle}
library(bundle)

bundled_model <- bundle(model)
saveRDS(bundled_model, "mtcars-ridge.bundle.rds")

bundled_model <- readRDS("mtcars-ridge.bundle.rds")
model <- unbundle(bundled_model)
```

For an nvForest-backed model, `device` chooses where the bundle will restore.
Use this when training a random forest on a GPU and deploying it on a CPU-only
host.

```{r bundle-nvforest}
forest <- cuda_ml_rand_forest(
  Species ~ .,
  data = iris,
  trees = 100,
  seed = 1
)

cpu_bundle <- bundle(forest, device = "cpu")
saveRDS(cpu_bundle, "forest-cpu.bundle.rds")
```

The target host must install the CPU inference backend before calling
`unbundle()`:

```{r unbundle-nvforest}
library(cuda.ml)
library(bundle)

cuda_ml_install(device = "cpu")
forest <- unbundle(readRDS("forest-cpu.bundle.rds"))
```

The `device` argument is supported only for nvForest-backed models.

## Export an nvForest checkpoint pair

`cuda_ml_nvforest_export()` writes two files:

- `<prefix>.treelite.checkpoint`, containing the device-neutral trees;
- `<prefix>.cuda-ml.json`, containing the metadata needed for an exact cuda.ml
  round-trip.

```{r export-nvforest}
dir.create("forest-artifact")
cuda_ml_nvforest_export(
  forest,
  directory = "forest-artifact",
  prefix = "model"
)
```

Copy both files when another cuda.ml process will restore the model. Select the
deployment device during import:

```{r import-nvforest}
cuda_ml_install(device = "cpu")

forest <- cuda_ml_nvforest_import(
  directory = "forest-artifact",
  prefix = "model",
  device = "cpu"
)
```

Other Treelite consumers can read the checkpoint alone, but they must supply
predictors in the recorded order and reproduce any class-label and
postprocessing semantics in the JSON sidecar. Loading the bare checkpoint back
into cuda.ml does not recover those semantics. Use the pair for an exact
round-trip.

## Compatibility rules

Every current cuda.ml state uses schema 1 and records its model ABI, package
version, backend provenance, and payload. The package version is provenance; a
different package version does not by itself prevent restoration. Compatibility
depends on the state schema, model ABI, and the native format used by the
payload.

| Model state | Required backend identity |
|:---|:---|
| Linear and logistic regression | No backend identity match; the state payload is portable R data |
| PCA, SVC, one-vs-rest SVC, SVR, and UMAP | Exact RAPIDS version |
| Random forest and nvForest | Exact Treelite version |

These model families implement explicit state in the current release. Models
without explicit state support fail during `cuda_ml_serialize()` rather than
falling back to serialization of native pointers.

States are not migrated implicitly. An unknown schema, unsupported model ABI,
missing payload, or missing or unequal required backend field produces an
error before restoration.

## Select the nvForest restore device

Current random-forest and nvForest states contain device-neutral Treelite model
bytes. Select CPU or GPU inference while restoring:

```{r nvforest-devices}
forest_state <- cuda_ml_serialize(forest)
cpu_forest <- cuda_ml_unserialize(forest_state, device = "cpu")
gpu_forest <- cuda_ml_unserialize(
  forest_state,
  device = "gpu",
  device_id = 0
)
```

If `device` is omitted, these states restore for GPU inference. The state keeps
prediction precision, class labels, preprocessing, and model semantics. It does
not keep the deployment device, device identifier, tree layout, chunk size, or
memory alignment. Restore-time inference options are supported only for current
nvForest and random-forest states.

Install the complete backend for GPU operation or the smaller CPU backend for
CPU-only nvForest inference. See the
[installation and runtime guide](install-manage.html) for those workflows.

## Treat model artifacts as trusted input

Load cuda.ml states, bundles, and checkpoint sidecars only from trusted
sources. An nvForest JSON sidecar embeds an R-serialized preprocessing
blueprint. Its SHA-256 digest checks checkpoint integrity, not the identity of
the artifact's producer.
