---
title: "nvForest inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nvForest inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

nvForest runs inference for tree ensembles exported by XGBoost, LightGBM, or
Treelite. The examples in this vignette are not evaluated when the vignette is
built. Building it therefore does not require model files, a GPU, the managed
runtime, network access, or a runtime download.

## Prepare the runtime

Install the native backend and its managed runtime before loading a model:

```{r install}
library(cuda.ml)
cuda_ml_install()
```

CPU inference does not require an NVIDIA GPU or driver. It still requires
`cuda_ml_install()` and the full, roughly 1.6 GiB managed CUDA/RAPIDS runtime.
Provision that runtime while preparing the deployment environment, not while
handling a prediction request. A completed managed cache is reused by later R
processes.

## Load exported models

`cuda_ml_nvforest_load_model()` accepts exactly five `model_type` values:

| `model_type` | Exported model format |
|:---|:---|
| `"xgboost_ubj"` | XGBoost UBJSON |
| `"xgboost_json"` | XGBoost JSON |
| `"xgboost_legacy"` | XGBoost legacy binary |
| `"lightgbm"` | LightGBM text |
| `"treelite_checkpoint"` | Treelite checkpoint |

The format can always be selected explicitly:

```{r explicit-formats}
xgb_ubjson <- cuda_ml_nvforest_load_model(
  "xgboost-model.ubj",
  model_type = "xgboost_ubj",
  device = "cpu"
)
xgb_json <- cuda_ml_nvforest_load_model(
  "xgboost-model.json",
  model_type = "xgboost_json",
  device = "cpu"
)
xgb_legacy <- cuda_ml_nvforest_load_model(
  "xgboost-model.model",
  model_type = "xgboost_legacy",
  device = "cpu"
)
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  model_type = "lightgbm",
  device = "cpu"
)
treelite_model <- cuda_ml_nvforest_load_model(
  "treelite-model.checkpoint",
  model_type = "treelite_checkpoint",
  device = "cpu"
)
```

With `model_type = NULL`, the default, the loader infers four formats from a
case-insensitive file suffix:

| Suffix | Inferred format |
|:---|:---|
| `.ubj` | XGBoost UBJSON |
| `.json` | XGBoost JSON |
| `.model` | XGBoost legacy binary |
| `.txt` | LightGBM text |

```{r inferred-formats}
xgb_model <- cuda_ml_nvforest_load_model("xgboost-model.ubj", device = "cpu")
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  class_levels = c("no", "yes"),
  device = "cpu"
)
```

Treelite checkpoint filenames are not inferred. Supply
`model_type = "treelite_checkpoint"` for every checkpoint filename.

## Select CPU or GPU inference

Select the device when loading the model. The default is GPU, so set the device
explicitly for CPU deployment.

```{r devices}
cpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "cpu"
)

gpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "gpu",
  device_id = 0
)
```

`device_id` applies only to GPU models. Omitting it uses the current CUDA
device. CPU and GPU models otherwise use the same prediction and inspection
APIs.

## Predict regression and classification results

Regression prediction returns a data frame with a `.pred` column:

```{r regression}
regression_model <- cuda_ml_nvforest_load_model(
  "regression.ubj",
  device = "cpu"
)

new_data <- data.frame(
  feature_1 = c(0.2, 0.8),
  feature_2 = c(1.5, 0.4)
)

regression_predictions <- predict(regression_model, new_data)
```

For a classifier, `class_levels` supplies labels in model-output order.
Class prediction returns `.pred_class`. Probability prediction returns one
`.pred_<level>` column for each class.

```{r classification}
classifier <- cuda_ml_nvforest_load_model(
  "classifier.txt",
  model_type = "lightgbm",
  class_levels = c("no", "yes"),
  device = "cpu"
)

class_predictions <- predict(classifier, new_data, type = "class")
probability_predictions <- predict(classifier, new_data, type = "prob")
```

Probability prediction is available only when the model's Treelite
postprocessor produces probabilities.

## Inspect a model

Use `cuda_ml_nvforest_info()` to inspect the task, model dimensions, output
representation, inference device, layout, precision, and chunk settings.

```{r inspect}
info <- cuda_ml_nvforest_info(classifier)
info$task_type
info$num_features
info$num_trees
info$device
info$has_probability_output
```

## Inspect leaves and individual trees

Leaf identifiers form an integer matrix with one row per observation and one
column per tree.

```{r leaves}
leaf_ids <- cuda_ml_nvforest_leaf_ids(classifier, new_data)
```

Individual-tree predictions form a numeric `rows × trees` matrix for
scalar-leaf models. Vector-leaf models return a numeric
`rows × trees × outputs` array.

```{r per-tree}
per_tree <- cuda_ml_nvforest_predict_per_tree(classifier, new_data)
```

Per-tree prediction fully materializes the logical
`rows × trees × outputs` result in R; scalar-leaf models omit the singleton
output dimension. `chunk_size` changes native inference batching, but it does
not bound the R result's memory. Account for the complete result when choosing
the number of rows and trees for this operation.

## Persist and restore a model

Use `cuda_ml_serialize()` rather than saving the live model object. It creates
an explicit schema 1 model state containing the nvForest payload and its
compatibility metadata.

```{r save-state}
state <- cuda_ml_serialize(classifier)
saveRDS(state, "classifier.cuda-ml-state.rds")
```

Restore the state with `cuda_ml_unserialize()` in a fresh R process. Prepare the
managed runtime on that host before restoration.

```{r restore-state}
library(cuda.ml)
cuda_ml_install()

state <- readRDS("classifier.cuda-ml-state.rds")
classifier <- cuda_ml_unserialize(state)
predict(classifier, new_data, type = "prob")
```

Schema 1 restoration requires the installed package to support the saved state
class and model ABI. An nvForest state contains serialized Treelite model bytes,
so its recorded `treelite_version` must exactly match the target backend.
The package version and other recorded backend fields are provenance rather
than compatibility gates for schema 1. States are not migrated implicitly, and
native pointers are not used as a persistence fallback.

## Serve predictions with Plumber

The following Plumber-annotated file restores one model when the serving process
starts and uses that model for every request. The request body is JSON with a
column-oriented `predictors` object, such as
`{"predictors":{"feature_1":[0.2],"feature_2":[1.5]}}`.

```{r plumber}
# plumber.R
library(cuda.ml)

cuda_ml_install()
classifier <- cuda_ml_unserialize(
  readRDS("classifier.cuda-ml-state.rds")
)

#* Return class probabilities
#* @post /predict
#* @serializer unboxedJSON
function(req) {
  new_data <- as.data.frame(req$body$predictors)
  predict(classifier, new_data, type = "prob")
}
```

The route uses only the public cuda.ml persistence and prediction APIs. Prepare
the managed cache before starting the service so `cuda_ml_install()` reuses the
completed installation without downloading files.
