---
title: "nvForest inference and deployment"
description: >
  Train random forests on a GPU, import tree ensembles, and deploy nvForest
  inference on a CPU or GPU.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nvForest inference and deployment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

nvForest is the inference layer for random forests trained by cuda.ml and for
tree ensembles exported by XGBoost, LightGBM, or Treelite. cuda.ml trains its
random forests with cuML on a GPU. The resulting model state is device neutral,
so it can be restored for GPU inference or deployed to a host without a GPU for
CPU inference. Imported XGBoost and LightGBM models can likewise run on either
device; cuda.ml does not train those model types.

For an externally trained XGBoost or LightGBM model, GPU inference is the
CUDA-accelerated use case. CPU mode is a deployment option that preserves the
same nvForest API on hosts without a GPU; it is not CUDA acceleration. The
GPU-training-to-CPU-deployment workflow supplied entirely by cuda.ml applies
to `cuda_ml_rand_forest()` models.

The examples in this vignette are not evaluated when the vignette is built.
Building it therefore does not require model files, a GPU, a native backend,
network access, or a runtime download.

## Prepare the runtime

Install the complete backend for cuML training or GPU inference. Install the
smaller CPU backend on a host used only for nvForest CPU inference.

```{r install}
library(cuda.ml)

# Complete backend for training and GPU inference.
cuda_ml_install()

# Smaller backend on a CPU-only inference host.
cuda_ml_install(device = "cpu")
```

Installation never occurs while loading or predicting. Provision the required
backend while preparing each environment. See the
[installation and runtime guide](install-manage.html) for system requirements,
runtime sizes, cache management, mirrors, and source builds.

## Train on GPU and deploy on CPU

`cuda_ml_rand_forest()` trains a cuML random forest on a GPU. Its fitted object
uses nvForest for inference and supports the same prediction and inspection
functions as imported models.

```{r train-random-forest}
library(cuda.ml)
cuda_ml_install()

set.seed(1)
forest <- cuda_ml_rand_forest(
  class ~ .,
  data = modeldata::hpc_data,
  trees = 500
)

dir.create("hpc-forest")
cuda_ml_nvforest_export(
  forest,
  directory = "hpc-forest",
  prefix = "model"
)
```

This writes `model.treelite.checkpoint` and `model.cuda-ml.json`. Copy both
files to a CPU host. That host needs only the CPU inference backend:

```{r deploy-random-forest}
library(cuda.ml)
cuda_ml_install(device = "cpu")

forest <- cuda_ml_nvforest_import(
  directory = "hpc-forest",
  prefix = "model",
  device = "cpu"
)

hpc_predictors <- modeldata::hpc_data[
  names(modeldata::hpc_data) != "class"
]
predict(forest, hpc_predictors[1:5, ], type = "class")
predict(forest, hpc_predictors[1:5, ], type = "prob")
```

The checkpoint is a standard, device-neutral Treelite checkpoint containing
the trees. The JSON is cuda.ml metadata, not an XGBoost-style model file. It
binds the pair with the checkpoint size and SHA-256 digest and retains class
labels, prediction precision,
random-forest probability semantics, R preprocessing blueprint, and processed
feature order when names are available. Exact cuda.ml round-trip requires both
files. Other Treelite consumers can load the checkpoint alone, but must supply
numeric predictors in the recorded processed order, or in the checkpoint's
original positional order when names are absent. They must also implement the
sidecar's labels and postprocessing semantics.

The sidecar does not encode the inference device, GPU identifier, tree layout,
chunk size, or memory alignment. Those are deployment choices. It embeds an
R-serialized preprocessing blueprint, which `cuda_ml_nvforest_import()`
unserializes. Import only artifacts from trusted sources. SHA-256 checks
integrity, not authenticity.

## Load external model formats

`cuda_ml_nvforest_load_model()` accepts exactly five `model_type` values:

| `model_type` | Exported model format |
|:---|:---|
| `"xgboost_ubj"` | XGBoost UBJSON |
| `"xgboost_json"` | XGBoost JSON |
| `"xgboost_legacy"` | XGBoost legacy binary |
| `"lightgbm"` | LightGBM text |
| `"treelite_checkpoint"` | Treelite checkpoint |

The format can always be selected explicitly:

```{r explicit-formats}
xgb_ubjson <- cuda_ml_nvforest_load_model(
  "xgboost-model.ubj",
  model_type = "xgboost_ubj",
  device = "cpu"
)
xgb_json <- cuda_ml_nvforest_load_model(
  "xgboost-model.json",
  model_type = "xgboost_json",
  device = "cpu"
)
xgb_legacy <- cuda_ml_nvforest_load_model(
  "xgboost-model.model",
  model_type = "xgboost_legacy",
  device = "cpu"
)
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  model_type = "lightgbm",
  device = "cpu"
)
treelite_model <- cuda_ml_nvforest_load_model(
  "treelite-model.checkpoint",
  model_type = "treelite_checkpoint",
  device = "cpu"
)
```

With `model_type = NULL`, the default, the loader infers four formats from a
case-insensitive file suffix:

| Suffix | Inferred format |
|:---|:---|
| `.ubj` | XGBoost UBJSON |
| `.json` | XGBoost JSON |
| `.model` | XGBoost legacy binary |
| `.txt` | LightGBM text |

```{r inferred-formats}
xgb_model <- cuda_ml_nvforest_load_model("xgboost-model.ubj", device = "cpu")
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  class_levels = c("no", "yes"),
  device = "cpu"
)
```

Treelite checkpoint filenames are not inferred. Supply
`model_type = "treelite_checkpoint"` for every checkpoint filename.

## Select CPU or GPU inference

Select the device when loading the model. The default is GPU, so set the device
explicitly for CPU deployment.

```{r devices}
cpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "cpu"
)

gpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "gpu",
  device_id = 0
)
```

`device_id` applies only to GPU models. Omitting it uses the current CUDA
device. CPU and GPU models otherwise use the same prediction and inspection
APIs.

## Predict regression and classification results

Models loaded directly from XGBoost, LightGBM, or Treelite files do not contain
a cuda.ml preprocessing blueprint or feature-name mapping. Supply numeric
predictors in exactly the column order used to train and export the model;
column names are not used to reorder them.

Regression prediction returns a data frame with a `.pred` column:

```{r regression}
regression_model <- cuda_ml_nvforest_load_model(
  "regression.ubj",
  device = "cpu"
)

new_data <- data.frame(
  feature_1 = c(0.2, 0.8),
  feature_2 = c(1.5, 0.4)
)

regression_predictions <- predict(regression_model, new_data)
```

For a classifier, `class_levels` supplies labels in model-output order.
Class prediction returns `.pred_class`. Probability prediction returns one
`.pred_<level>` column for each class.

```{r classification}
classifier <- cuda_ml_nvforest_load_model(
  "classifier.txt",
  model_type = "lightgbm",
  class_levels = c("no", "yes"),
  device = "cpu"
)

class_predictions <- predict(classifier, new_data, type = "class")
probability_predictions <- predict(classifier, new_data, type = "prob")
```

Probability prediction is available only when the model's Treelite
postprocessor produces probabilities.

## Serve predictions with plumber2

The following plumber2 route file restores one model when the serving process
starts and uses that model for every request. The request body is JSON with a
column-oriented `predictors` object, such as
`{"predictors":{"feature_1":[0.2],"feature_2":[1.5]}}`.

```r
# plumber2.R
library(cuda.ml)

classifier <- cuda_ml_nvforest_import(
  directory = "models",
  prefix = "classifier",
  device = "cpu"
)

#* Return class probabilities
#*
#* @post /predict
#* @parser json
#* @serializer unboxedJSON
function(body) {
  new_data <- as.data.frame(body$predictors)
  predict(classifier, new_data, type = "prob")
}
```

Start the service with plumber2 after provisioning the CPU inference backend:

```r
plumber2::api("plumber2.R") |>
  plumber2::api_run(host = "0.0.0.0", port = 8000)
```

The route uses only public cuda.ml persistence and prediction APIs. Run the
installer while building or provisioning the service image so startup and
requests need no network access.

## Inspect a model

Use `cuda_ml_nvforest_info()` to inspect the task, model dimensions, output
representation, inference device, layout, precision, and chunk settings.

```{r inspect}
info <- cuda_ml_nvforest_info(classifier)
info$task_type
info$num_features
info$num_trees
info$device
info$has_probability_output
```

## Inspect leaves and individual trees

Leaf identifiers form an integer matrix with one row per observation and one
column per tree.

```{r leaves}
leaf_ids <- cuda_ml_nvforest_leaf_ids(classifier, new_data)
```

Individual-tree predictions form a numeric `rows × trees` matrix for
scalar-leaf models. Vector-leaf models return a numeric
`rows × trees × outputs` array.

```{r per-tree}
per_tree <- cuda_ml_nvforest_predict_per_tree(classifier, new_data)
```

Per-tree prediction fully materializes the logical
`rows × trees × outputs` result in R; scalar-leaf models omit the singleton
output dimension. `chunk_size` changes native inference batching, but it does
not bound the R result's memory. Account for the complete result when choosing
the number of rows and trees for this operation.

## Persist as a single R state

Use `cuda_ml_serialize()` and `cuda_ml_unserialize()` when a single R-native
artifact is preferable. Use `cuda_ml_nvforest_export()` and
`cuda_ml_nvforest_import()` when the Treelite checkpoint must also be
available independently. Do not save the live model object.

```{r save-state}
cuda_ml_serialize(classifier, "classifier.cuda-ml-state")
classifier <- cuda_ml_unserialize(
  "classifier.cuda-ml-state",
  device = "cpu"
)
```

Current nvForest and random-forest states are device neutral. Select the
deployment device while restoring and prepare that backend first; cuda.ml
checks it before loading the model. The
[model persistence guide](model-persistence.html) compares state, bundle, and
checkpoint workflows.
