---
title: "nvForest inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nvForest inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

nvForest is the inference layer for random forests trained by cuda.ml and for
tree ensembles exported by XGBoost, LightGBM, or Treelite. cuda.ml trains its
random forests with cuML on a GPU. The resulting model state is device neutral,
so it can be restored for GPU inference or deployed to a host without a GPU for
CPU inference. Imported XGBoost and LightGBM models can likewise run on either
device; cuda.ml does not train those model types.

For an externally trained XGBoost or LightGBM model, GPU inference is the
CUDA-accelerated use case. CPU mode is a deployment option that preserves the
same nvForest API on hosts without a GPU; it is not CUDA acceleration. The
GPU-training-to-CPU-deployment workflow supplied entirely by cuda.ml applies
to `cuda_ml_rand_forest()` models.

The examples in this vignette are not evaluated when the vignette is built.
Building it therefore does not require model files, a GPU, a native backend,
network access, or a runtime download.

## Prepare the runtime

Install the complete backend before training with cuML or running GPU
inference:

```{r install}
library(cuda.ml)
cuda_ml_install()
```

That installation is roughly 1.6 GiB and contains the managed CUDA, RAPIDS
cuML, and nvForest runtime. For CPU-only nvForest inference, install the
separate backend instead:

```{r install-cpu}
cuda_ml_install(device = "cpu")
```

The CPU backend is roughly 1 MiB to download and 3 MiB when installed. It
does not include cuML, CUDA runtime libraries, or the complete managed
CUDA/RAPIDS runtime. It requires neither an NVIDIA GPU nor an NVIDIA driver,
and cuda.ml selects nvForest's CPU execution path.

An environment that already has the complete backend can continue to execute
nvForest models on CPU. The separate backend removes the complete runtime from
CPU-only deployments.

Installation never occurs while loading or predicting. Provision the required
backend while preparing each environment. A completed cache is reused by later
R processes.

## Train on GPU and deploy on CPU

`cuda_ml_rand_forest()` trains a cuML random forest on a GPU. Its fitted object
uses nvForest for inference and supports the same prediction and inspection
functions as imported models.

```{r train-random-forest}
library(cuda.ml)
cuda_ml_install()

forest <- cuda_ml_rand_forest(
  Species ~ .,
  data = iris,
  trees = 500,
  seed = 1
)

state <- cuda_ml_serialize(forest)
saveRDS(state, "iris-forest.cuda-ml-state.rds")
```

Copy the RDS file to a CPU host. That host needs only the CPU inference
backend:

```{r deploy-random-forest}
library(cuda.ml)
cuda_ml_install(device = "cpu")

forest <- cuda_ml_unserialize(
  readRDS("iris-forest.cuda-ml-state.rds"),
  device = "cpu"
)

predict(forest, iris[1:5, -5], type = "class")
predict(forest, iris[1:5, -5], type = "prob")
```

The persisted state contains the Treelite model representation, preprocessing
blueprint, class labels, and prediction semantics. It does not encode the
inference device, GPU identifier, tree layout, chunk size, or memory alignment.
Those are deployment choices. No bundle object is needed for this workflow;
the value returned by `cuda_ml_serialize()` is the deployment artifact.
If an existing workflow uses `bundle`, `bundle::bundle(forest, device = "cpu")`
binds CPU deployment to that bundle while leaving its nested model state
device neutral.

## Load exported models

`cuda_ml_nvforest_load_model()` accepts exactly five `model_type` values:

| `model_type` | Exported model format |
|:---|:---|
| `"xgboost_ubj"` | XGBoost UBJSON |
| `"xgboost_json"` | XGBoost JSON |
| `"xgboost_legacy"` | XGBoost legacy binary |
| `"lightgbm"` | LightGBM text |
| `"treelite_checkpoint"` | Treelite checkpoint |

The format can always be selected explicitly:

```{r explicit-formats}
xgb_ubjson <- cuda_ml_nvforest_load_model(
  "xgboost-model.ubj",
  model_type = "xgboost_ubj",
  device = "cpu"
)
xgb_json <- cuda_ml_nvforest_load_model(
  "xgboost-model.json",
  model_type = "xgboost_json",
  device = "cpu"
)
xgb_legacy <- cuda_ml_nvforest_load_model(
  "xgboost-model.model",
  model_type = "xgboost_legacy",
  device = "cpu"
)
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  model_type = "lightgbm",
  device = "cpu"
)
treelite_model <- cuda_ml_nvforest_load_model(
  "treelite-model.checkpoint",
  model_type = "treelite_checkpoint",
  device = "cpu"
)
```

With `model_type = NULL`, the default, the loader infers four formats from a
case-insensitive file suffix:

| Suffix | Inferred format |
|:---|:---|
| `.ubj` | XGBoost UBJSON |
| `.json` | XGBoost JSON |
| `.model` | XGBoost legacy binary |
| `.txt` | LightGBM text |

```{r inferred-formats}
xgb_model <- cuda_ml_nvforest_load_model("xgboost-model.ubj", device = "cpu")
lightgbm_model <- cuda_ml_nvforest_load_model(
  "lightgbm-model.txt",
  class_levels = c("no", "yes"),
  device = "cpu"
)
```

Treelite checkpoint filenames are not inferred. Supply
`model_type = "treelite_checkpoint"` for every checkpoint filename.

## Select CPU or GPU inference

Select the device when loading the model. The default is GPU, so set the device
explicitly for CPU deployment.

```{r devices}
cpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "cpu"
)

gpu_model <- cuda_ml_nvforest_load_model(
  "model.ubj",
  device = "gpu",
  device_id = 0
)
```

`device_id` applies only to GPU models. Omitting it uses the current CUDA
device. CPU and GPU models otherwise use the same prediction and inspection
APIs.

## Predict regression and classification results

Regression prediction returns a data frame with a `.pred` column:

```{r regression}
regression_model <- cuda_ml_nvforest_load_model(
  "regression.ubj",
  device = "cpu"
)

new_data <- data.frame(
  feature_1 = c(0.2, 0.8),
  feature_2 = c(1.5, 0.4)
)

regression_predictions <- predict(regression_model, new_data)
```

For a classifier, `class_levels` supplies labels in model-output order.
Class prediction returns `.pred_class`. Probability prediction returns one
`.pred_<level>` column for each class.

```{r classification}
classifier <- cuda_ml_nvforest_load_model(
  "classifier.txt",
  model_type = "lightgbm",
  class_levels = c("no", "yes"),
  device = "cpu"
)

class_predictions <- predict(classifier, new_data, type = "class")
probability_predictions <- predict(classifier, new_data, type = "prob")
```

Probability prediction is available only when the model's Treelite
postprocessor produces probabilities.

## Inspect a model

Use `cuda_ml_nvforest_info()` to inspect the task, model dimensions, output
representation, inference device, layout, precision, and chunk settings.

```{r inspect}
info <- cuda_ml_nvforest_info(classifier)
info$task_type
info$num_features
info$num_trees
info$device
info$has_probability_output
```

## Inspect leaves and individual trees

Leaf identifiers form an integer matrix with one row per observation and one
column per tree.

```{r leaves}
leaf_ids <- cuda_ml_nvforest_leaf_ids(classifier, new_data)
```

Individual-tree predictions form a numeric `rows × trees` matrix for
scalar-leaf models. Vector-leaf models return a numeric
`rows × trees × outputs` array.

```{r per-tree}
per_tree <- cuda_ml_nvforest_predict_per_tree(classifier, new_data)
```

Per-tree prediction fully materializes the logical
`rows × trees × outputs` result in R; scalar-leaf models omit the singleton
output dimension. `chunk_size` changes native inference batching, but it does
not bound the R result's memory. Account for the complete result when choosing
the number of rows and trees for this operation.

## Persist and restore a model

Use `cuda_ml_serialize()` rather than saving the live model object. It creates
an explicit schema 1 model state containing the nvForest payload and its
compatibility metadata.

```{r save-state}
state <- cuda_ml_serialize(classifier)
saveRDS(state, "classifier.cuda-ml-state.rds")
```

Restore the state with `cuda_ml_unserialize()` in a fresh R process. Select the
deployment device during restoration and prepare its backend first.

```{r restore-state}
library(cuda.ml)
cuda_ml_install(device = "cpu")

state <- readRDS("classifier.cuda-ml-state.rds")
classifier <- cuda_ml_unserialize(state, device = "cpu")
predict(classifier, new_data, type = "prob")
```

Schema 1 restoration requires the installed package to support the saved state
class and model ABI. Current nvForest and random-forest states are device
neutral. They retain prediction precision unless it is overridden during
restoration; other device-specific settings use nvForest defaults unless they
are supplied to `cuda_ml_unserialize()`. An nvForest state contains serialized
Treelite model bytes, so its recorded `treelite_version` must exactly match the
target backend. The package version and other recorded backend fields are
provenance rather than compatibility gates for schema 1. States are not
migrated implicitly, and native pointers are not used as a persistence
fallback. Legacy nvForest states retain their recorded inference settings.

## Serve predictions with Plumber

The following Plumber-annotated file restores one model when the serving process
starts and uses that model for every request. The request body is JSON with a
column-oriented `predictors` object, such as
`{"predictors":{"feature_1":[0.2],"feature_2":[1.5]}}`.

```{r plumber}
# plumber.R
library(cuda.ml)

cuda_ml_install(device = "cpu")
classifier <- cuda_ml_unserialize(
  readRDS("classifier.cuda-ml-state.rds"),
  device = "cpu"
)

#* Return class probabilities
#* @post /predict
#* @serializer unboxedJSON
function(req) {
  new_data <- as.data.frame(req$body$predictors)
  predict(classifier, new_data, type = "prob")
}
```

The route uses only public cuda.ml persistence and prediction APIs. Run the
installer while building or provisioning the service image, before starting
Plumber, so service startup and requests need no network access.
