---
title: "Use cuda.ml with tidymodels"
description: >
  Fit cuda.ml models through parsnip, preprocess predictors with recipes, and
  distinguish portable model arguments from engine-specific controls.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use cuda.ml with tidymodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  eval = identical(Sys.getenv("IN_PKGDOWN"), "true") ||
    identical(Sys.getenv("CUDA_ML_GPU_VIGNETTES"), "true")
)
```

cuda.ml registers parsnip engines for its supervised models. This gives the
models the same specification, fitting, and prediction interface as other
parsnip engines while training with cuML. Use the direct cuda.ml API when you
need an algorithm that has no parsnip specification or detailed control over
a solver. See [Getting started](cuda-ml.html) for installation and runtime
setup.

## Available model specifications

Loading cuda.ml registers the engine name `"cuda.ml"` for these parsnip
specifications:

| parsnip specification | mode | prediction types |
|:--|:--|:--|
| `linear_reg()` | regression | `"numeric"` |
| `logistic_reg()` | classification | `"class"`, `"prob"` |
| `multinom_reg()` | classification | `"class"`, `"prob"` |
| `rand_forest()` | classification | `"class"`, `"prob"` |
| `rand_forest()` | regression | `"numeric"` |
| `nearest_neighbor()` | classification | `"class"`, `"prob"` |
| `nearest_neighbor()` | regression | `"numeric"` |
| `svm_rbf()` | classification | `"class"` |
| `svm_rbf()` | regression | `"numeric"` |
| `svm_poly()` | classification | `"class"` |
| `svm_poly()` | regression | `"numeric"` |
| `svm_linear()` | classification | `"class"` |
| `svm_linear()` | regression | `"numeric"` |

The SVM engines do not register probability predictions. In particular,
`predict(fitted_svm, new_data, type = "prob")` is not supported. Use
`logistic_reg()`, `multinom_reg()`, `rand_forest()`, or
`nearest_neighbor()` when a classification workflow requires probabilities.

Use `logistic_reg()` for a two-level outcome and `multinom_reg()` for an
outcome with more than two levels. Both specifications call
`cuda_ml_logistic_reg()`; the outcome determines whether cuda.ml uses its
binary or multinomial loss.

## Set the engine

Set the mode and common arguments in a parsnip specification, then select the
cuda.ml engine:

```{r random-forest-specification}
library(cuda.ml)
library(parsnip)

forest_spec <- rand_forest(
  mode = "classification",
  mtry = 2,
  trees = 500,
  min_n = 5
) |>
  set_engine(
    "cuda.ml",
    max_depth = 20L,
    n_bins = 256L,
    seed = 1L
  )

forest_fit <- fit(forest_spec, class ~ ., data = modeldata::hpc_data)

class_predictions <- predict(forest_fit, modeldata::hpc_data, type = "class")
probabilities <- predict(forest_fit, modeldata::hpc_data, type = "prob")
```

Arguments in the model specification, such as `mtry`, are common parsnip
arguments. Arguments in `set_engine()`, such as `max_depth`, are specific to
cuda.ml. Put each argument in only one place.

## Preprocess with a recipe

cuda.ml's supervised models require numeric predictors. Scaling is especially
important for KNN and SVM models because their fits depend on distances or
margins. It is also usually appropriate for penalized linear and logistic
models. Fit preprocessing parameters on the training data only, then apply
the same recipe to assessment or production data.

This example normalizes the predictors before fitting an exact KNN
classifier. A workflow keeps the preprocessing and model specifications
together. `fit()` estimates the recipe from the training data before fitting
the cuda.ml model, and `predict()` applies the recipe to new data before
calling the model engine.

```{r recipe-example}
library(cuda.ml)
library(parsnip)
library(recipes)
library(workflows)

set.seed(1)
training_rows <- sample(
  seq_len(nrow(modeldata::two_class_dat)),
  floor(0.8 * nrow(modeldata::two_class_dat))
)
training_data <- modeldata::two_class_dat[training_rows, ]
testing_data <- modeldata::two_class_dat[-training_rows, ]

classifier_recipe <- recipe(Class ~ ., data = training_data) |>
  step_normalize(all_numeric_predictors())

knn_spec <- nearest_neighbor(
  mode = "classification",
  neighbors = 5,
  dist_power = 2
) |>
  set_engine(
    "cuda.ml",
    algo = "brute",
    metric = "euclidean"
  )

knn_workflow <- workflow() |>
  add_recipe(classifier_recipe) |>
  add_model(knn_spec)

knn_fit <- fit(knn_workflow, data = training_data)

results <- cbind(
  truth = testing_data$Class,
  predict(knn_fit, testing_data, type = "class"),
  predict(knn_fit, testing_data, type = "prob")
)
head(results)
```

The parsnip KNN engine defaults to `algo = "ivfflat"` and
`metric = "euclidean"`. `ivfflat` performs approximate neighbor search. Set
`algo = "brute"`, as above, when exact search is required. The direct
`cuda_ml_knn()` interface defaults to brute-force search. The cuda.ml engine
does not map parsnip's `weight_func` argument; leave it as `NULL`.

Tree models do not generally need normalization. A recipe can still be useful
for creating numeric indicators or applying other preprocessing learned from
the training set.

## Tune common arguments

cuda.ml registers standard dials parameter metadata for the following parsnip
arguments:

| specification | registered arguments |
|:--|:--|
| `linear_reg()` | `penalty`, `mixture` |
| `logistic_reg()`, `multinom_reg()` | `penalty`, `mixture` |
| `rand_forest()` | `mtry`, `trees`, `min_n` |
| `nearest_neighbor()` | `neighbors`, `dist_power` |
| `svm_rbf()` | `cost`, `margin`, `rbf_sigma` |
| `svm_poly()` | `cost`, `margin`, `degree`, `scale_factor` |
| `svm_linear()` | `cost`, `margin` |

These arguments can use `tune::tune()` in a tuning workflow. `margin` maps to
the epsilon tube and affects SVM regression only; do not tune it for an SVM
classifier.

Engine arguments are not registered as dials parameters. Keep them fixed in
`set_engine()` unless you define an explicit dials parameter and range for
the tuning workflow. Useful engine arguments include:

| engine | examples of engine-specific arguments |
|:--|:--|
| linear regression | `fit_intercept`; route-specific options described below |
| logistic and multinomial regression | `fit_intercept`, `tol`, `class_weight`, `max_iter`, `linesearch_max_iter`, `lbfgs_memory`, `penalty_normalized` |
| random forest | `bootstrap`, `sample_fraction`, `max_depth`, `max_leaves`, `n_bins`, `min_samples_leaf`, `split_criterion`, `min_impurity_decrease`, `max_batch_size`, `n_streams`, `seed` |
| nearest neighbor | `algo`, `metric` |
| SVM | `coef0`, `tol`, `max_iter`, `nochange_steps`, `cache_size` |

Parsnip case weights are not currently supported by the cuda.ml engine. For
per-observation weights, use `sample_weight` with `cuda_ml_logistic_reg()` or
`sample_weights` with `cuda_ml_svm()` directly.

Consult the corresponding `cuda_ml_*()` reference page before setting these
arguments. For example, random-forest split criteria differ between
classification and regression, and approximate KNN algorithms support fewer
distance metrics than brute-force KNN.

### Linear regression routing

The `linear_reg()` engine selects a cuda.ml solver from `penalty` and
`mixture`:

| values | direct function |
|:--|:--|
| `penalty = NULL` or `penalty = 0` | `cuda_ml_ols()` |
| positive `penalty`, `mixture = 0` | `cuda_ml_ridge()` |
| positive `penalty`, `mixture = 1` or `NULL` | `cuda_ml_lasso()` |
| positive `penalty`, `0 < mixture < 1` | `cuda_ml_elastic_net()` |

Only pass options supported by the selected function. For example, `method`
is an OLS option, while `max_iter`, `tol`, and `selection` apply to the lasso
and elastic-net routes. If the solver itself is part of the decision, use the
named direct functions so that the relationship between the function and its
arguments remains explicit.

## Choose between parsnip and the direct API

Use parsnip when you want to compare engines through common specifications,
use a tidymodels tuning workflow, or consume standard parsnip prediction
types. Parsnip delegates training and prediction to cuda.ml's public model
functions.

The direct API includes several capabilities that have no parsnip
specification:

| Capability | parsnip | Direct cuda.ml API |
|:--|:--|:--|
| Linear, logistic, random-forest, KNN, and RBF, polynomial, or linear SVM models | Supported | Supported |
| Clustering and dimensionality reduction | Not available | Agglomerative clustering, DBSCAN, k-means, PCA, tSVD, UMAP, and t-SNE |
| Transform and inverse-transform operations | Not available | `cuda_ml_transform()` and `cuda_ml_inverse_transform()` |
| Stochastic-gradient-descent linear regression | No dedicated route | `cuda_ml_sgd()` |
| Hyperbolic-tangent SVM | No matching specification | `cuda_ml_svm(kernel = "tanh")` |
| External tree-ensemble inference and inspection | Not available | nvForest load, predict, model-info, leaf-ID, per-tree, import, and export functions |

Most algorithm-specific arguments for supported parsnip models remain
available through `set_engine()`. Use the named direct function when selecting
the algorithm or solver is part of the analysis, or when passing predictors and
outcomes directly is preferable to a parsnip workflow.

For example, cuda.ml's direct SVM interface supports a `"tanh"` kernel, but
parsnip registration is limited to the RBF, polynomial, and linear SVM
specifications.

```{r direct-api}
direct_fit <- cuda_ml_svm(
  Class ~ .,
  data = modeldata::two_class_dat,
  kernel = "tanh",
  cost = 2,
  gamma = 0.1,
  coef0 = 0
)

direct_predictions <- predict(
  direct_fit,
  modeldata::two_class_dat[c("A", "B")]
)
```

See [Save and restore models](model-persistence.html) before moving a fitted
model to another R process or deployment host.
